cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-17 paper-cardinalR.Rmd
Updated wb.ratio

Figure \@ref(fig:fig-cluster-stats) shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a $4-5$ cluster solution. Pearson gamma (`pearsongamma`) increases sharply up to five clusters before levelling off, Calinski–Harabasz index (`ch`) increases sharply from 4 to 5 clusters and Dunn (`dunn2`) has a maximum at 5 for two methods and at 4 for $k$-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (`wbratio`) and within-cluster sum of squares (`within.cluster.ss`) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (`sindex`) is optimised at 4 clusters for $k$-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, $k$-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

>>>>

Figure \@ref(fig:fig-cluster-stats) shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a $4-5$ cluster solution. Pearson gamma (`pearsongamma`) increases sharply up to five clusters before levelling off, Calinski–Harabasz index (`ch`) increases sharply from 4 to 5 clusters and Dunn (`dunn2`) has a maximum at 5 for two methods and at 4 for $k$-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (`wb.ratio`) and within-cluster sum of squares (`within.cluster.ss`) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (`sindex`) is optimised at 4 clusters for $k$-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, $k$-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

### 2025-12-17 paper-cardinalR.tex
Updated wb.ratio

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. Pearson gamma (\texttt{pearsongamma}) increases sharply up to five clusters before levelling off, Calinski--Harabasz index (\texttt{ch}) increases sharply from 4 to 5 clusters and Dunn (\texttt{dunn2}) has a maximum at 5 for two methods and at 4 for \(k\)-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (\texttt{wbratio}) and within-cluster sum of squares (\texttt{within.cluster.ss}) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (\texttt{sindex}) is optimised at 4 clusters for \(k\)-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, \(k\)-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

>>>>

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. Pearson gamma (\texttt{pearsongamma}) increases sharply up to five clusters before levelling off, Calinski--Harabasz index (\texttt{ch}) increases sharply from 4 to 5 clusters and Dunn (\texttt{dunn2}) has a maximum at 5 for two methods and at 4 for \(k\)-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (\texttt{wb.ratio}) and within-cluster sum of squares (\texttt{within.cluster.ss}) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (\texttt{sindex}) is optimised at 4 clusters for \(k\)-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, \(k\)-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.
